---
title: "week2"
author: "Roxana Niksefat"
date: "2023-11-06"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Health and Economic Impact of Weather Events in the US
======================================================


Adverse weather conditions, including storms, can lead to significant challenges for communities and local economies by affecting public health and causing economic repercussions. These events often result in fatalities, injuries, and property damage, making it crucial to prioritize measures that can help mitigate these outcomes.

The task at hand entails an investigation of the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This repository documents the specifics of major storms and weather occurrences in the United States, detailing their timing, location, and providing estimates of associated fatalities, injuries, and property damage.

Synopsis
========

Examination of the storm event database uncovered that tornadoes pose the greatest risk to public health among all weather events. Following closely as the second most perilous event type is excessive heat. Furthermore, the economic repercussions of these weather occurrences were assessed. Flash floods and thunderstorm winds resulted in billions of dollars in property damages from 1950 to 2011. The most substantial crop damage was attributed to drought, with flood and hail following in impact.


Data Processing
===============

The analysis was performed on
[Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp), provided by
[National Climatic Data Center](http://www.ncdc.noaa.gov/). The data is from a comma-separated-value file available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).
There is also some documentation of the data available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

The first step is to read the data into a data frame.

```{r}
storm <- read.csv(("repdata_data_StormData.csv"),header = TRUE)
```

Before the analysis, the data need some preprocessing. Event types don't have a
specific format. For instance, there are events with types `Frost/Freeze`,
`FROST/FREEZE` and `FROST\\FREEZE` which obviously refer to the same type of
event.



```{r}
# number of unique event types
length(unique(storm$EVTYPE))
```


```{r}
# translate all letters to lowercase
event_types <- tolower(storm$EVTYPE)
# replace all punct. characters with a space
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
length(unique(event_types))
```


```{r}
# update the data frame
storm$EVTYPE <- event_types
```

No further data preprocessing was performed although the event type field can be
processed further to merge event types such as `tstm wind` and `thunderstorm wind`. 
After the cleaning, as expected, the number of unique event types reduce
significantly. For further analysis, the cleaned event types are used.


Dangerous Events with respect to Population Health
================================================

To find the event types that are most harmful to population health, the number
of casualties are aggregated by the event type.


```{r}
library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))

# Find events that caused most death and injury
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
```

Top 10 events that caused largest number of deaths are


```{r}
fatal_events[, c("EVTYPE", "fatalities")]
```

##             EVTYPE fatalities
## 737        tornado       5633
## 109 excessive heat       1903
## 132    flash flood        978
## 234           heat        937
## 400      lightning        816
## 760      tstm wind        504
## 148          flood        470
## 511    rip current        368
## 309      high wind        248
## 11       avalanche        224

Top 10 events that caused most number of injuries are


```{r}
injury_events[, c("EVTYPE", "injuries")]
```


##                EVTYPE injuries
## 737           tornado    91346
## 760         tstm wind     6957
## 148             flood     6789
## 109    excessive heat     6525
## 400         lightning     5230
## 234              heat     2100
## 377         ice storm     1975
## 132       flash flood     1777
## 670 thunderstorm wind     1488
## 203              hail     1361

Economic Effects of Weather Events
==================================

To analyze the impact of weather events on the economy, available property
damage and crop damage reportings/estimates were used.

In the raw data, the property damage is represented with two fields, a number
`PROPDMG` in dollars and the exponent `PROPDMGEXP`. Similarly, the crop damage
is represented using two fields, `CROPDMG` and `CROPDMGEXP`. The first step in the
analysis is to calculate the property and crop damage for each event.


```{r}
exp_transform <- function(e) {
    # h -> hundred, k -> thousand, m -> million, b -> billion
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) # if a digit
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid exponent value.")
    }
}
```


```{r warning=FALSE}
prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)
```



```{r}
# Compute the economic loss by event type
library(plyr)
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))

# filter out events that caused no economic loss
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
```

Top 10 events that caused most property damage (in dollars) are as follows


```{r}
prop_dmg_events[, c("EVTYPE", "prop_dmg")]
```


Similarly, the events that caused biggest crop damage are


```{r}
crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```



Results
=======

Health impact of weather events
-------------------------------

The following plot shows top dangerous weather event types.


```{r mio}
library(ggplot2)
library(gridExtra)
# Set the levels in order
p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Total number of fatalities") +
    xlab("Event type") +
    theme(legend.position="none")

p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Total number of injuries") +
    xlab("Event type") +
    theme(legend.position="none")

grid.arrange(p1, p2)
```


Tornadoes have been accountable for the highest number of fatalities and injuries compared to all other types of weather events. Over the past 60 years in the United States, tornadoes have resulted in more than 5,000 deaths and over 10,000 injuries. Following closely in terms of endangering public health are excessive heat and flash floods among the most hazardous event types.

Economic impact of weather events
---------------------------------

The following plot shows the most severe weather event types with respect to
economic cost that they have costed since 1950s.



```{r plot}
library(ggplot2)
library(gridExtra)
# Set the levels in order
p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Event type") +
    ylab("Property damage in dollars (log-scale)") +
    theme(legend.position="none")

p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Event type") +
    ylab("Crop damage in dollars") + 
    theme(legend.position="none")

grid.arrange(p1, p2)
```


Property damages are given in logarithmic scale due to large range of values.
The data shows that flash floods and thunderstorm winds cost the largest
property damages among weather-related natural diseasters. Note that, due to
untidy nature of the available data, type `flood` and `flash flood` are
separate values and should be merged for more accurate data-driven conclusions.

The most severe weather event in terms of crop damage is the drought. In the last
half century, the drought has caused more than 10 billion dollars damage. Other
severe crop-damage-causing event types are floods and hails.
